---
title: "J1798 PDAC trial - CyTOF myeloid panel visualization"
author: "Ludmila Danilova"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      warning = FALSE)
rm(list = ls())

library(dplyr)
library(ggplot2)
library(ggpubr)
library(ggrepel)
library(data.table)
library("DT")
library(gridExtra)
library(scales)
library(readxl)
library(xlsx)
library(pheatmap)
library(pals)
library(kableExtra)

# set root folder
rootDir = "../"

# load shared parameters and functions
source(paste0(rootDir,"parameters_and_functions.r"))

```
Create figures for CyTOF myeloid panel data. Preprocessing and preliminary analysis were done by Dimitrios N. Sidiropoulos.

# Cell proportions

## Bar plot

```{r echo=FALSE}

#=========================
# read in sample annotation and mapping
patTab = read.csv(file = paste0(rootDir,"input_data/table_patients.csv"), row.names = 1)
# switch to R/NR annotation
patTab = patTab %>% mutate(RECIST = factor(RECIST,levels = c("Responder", "NonResponder"), labels = c("R","NR")))

#=============================
# read in proportions
props = read.csv(file = paste0(rootDir,"input_data/table_cytof_myeloid_props.csv"), row.names = 1)

# transpose, extract time point and add response
props = t(props)
props = data.frame(
  patientID = paste(sapply(strsplit(rownames(props), '_'), geti,1), sapply(strsplit(rownames(props), '_'), geti,2), sep = '_'),
  timepoint = sapply(strsplit(rownames(props), '_'), geti,3), 
  props, check.names = F)

table(props$timepoint)
# bar plot by time point
clusters = factor(colnames(props)[3:ncol(props)])

# add response
props = props %>% mutate(RECIST = patTab[patientID,"RECIST"])


# cluster colors
clustercolors <- toupper(as.character(c(cols25(n=25),alphabet(n=19)))[1:length(clusters)])
color_clusters1 <- clustercolors[1:nlevels(clusters)]
names(color_clusters1) <- levels(clusters)
# Stacked bar plot
df_melt = melt(props, id.vars = c("patientID", "timepoint"), measure.vars = clusters)
print(ggplot(df_melt, aes(fill=variable, y=as.numeric(value), x=timepoint)) + 
    geom_bar(position="fill", stat="identity")+
      scale_fill_manual(values=color_clusters1)+
      ylab("Cell proportions") + xlab("Time point"))
```

## Table

P-values from paired two-sided Wilcoxon test between time points and mean of difference between time points

```{r table, echo=FALSE}
# paired Wilcoxon

res = t(sapply(as.character(clusters), pairedWilcox, props))

res = round(res,3)
write.csv(cbind(cluster = rownames(res),res), row.names = F, file = paste0(rootDir,"output_data/table_cytof_myeloid_wilcox_pvalues_proportions.csv"))

knitr::kable(res, "simple",digits = 3)
# significant proportions
sigProp = apply(res[,grep("pValue",colnames(res))],2,function(x) names(x)[which(x < pVal)])

print("Significant change (p-value < 0.05) in proportions of cells in the following cell types:")
print(sigProp)

```
## Boxplots

For significant cell types only

```{r boxplots, echo=FALSE, fig.width= 4, fig.height= 3 }
sigClusters = unique(unlist(sigProp))

for(i in sigClusters)
{
  df = cbind(cluster = props[,i], props[,c("patientID", "timepoint", "RECIST") ] )
#print(i)
    print(ggpaired(df , x="timepoint",y="cluster", id = "patientID", line.size = .7, line.color = "RECIST",palette =  recistCol,
     title = i, xlab="Time point", ylab = "Cell proportions") +
       theme(legend.position="right") +
      scale_x_discrete(labels=timeLab))
      
}


```

# Average expression of markers by cluster

## Table

```{r echo=FALSE, fig.width= 4, fig.height= 3 }
# read in expression
expr = read.csv(file = paste0(rootDir,"input_data/table_cytof_myeloid_marker_expression.csv"))

# for every cluster run paired wilcoxon between time points for marker expression
markers = colnames(expr)[4:ncol(expr)]

# add response
expr = expr %>% mutate(RECIST = patTab[patientID,"RECIST"])

res1 = data.frame()
for(i in clusters)
{
  res1 = rbind(res1, 
              cbind(marker = markers, cluster = rep(i,length(markers)),
                    round(t(sapply(as.character(markers), pairedWilcox, expr %>% filter(cluster == i))),3)))
  
}

write.csv(res1, row.names = F, file = paste0(rootDir,"output_data/table_cytof_myeloid_wilcox_pvalues_marker_expression.csv"))

datatable(res1)

```

## Boxplots
### two time points

```{r two-time-points, echo=FALSE, fig.width= 4, fig.height= 3 }
sigMarkers = res1 %>% filter(pValue_BvsC1D1 < pVal)
print("The number of pairs (cluster, marker) with significant change (p-value < 0.05) in marker expression between baseline and C1D1:")
print(nrow(sigMarkers))

for(i in 1:nrow(sigMarkers))
{
  # marker
  m = sigMarkers[i,"marker"]
  # cluster
  k = sigMarkers[i,"cluster"]
  d = expr %>% filter(cluster == k)
  df = cbind(protein = d[,m], d[,c("patientID", "timepoint", "RECIST") ] )
#print(i)
  # remove C2D1 timepoint
  df = df %>% filter(timepoint != "C2D1")
  
  print(paste0(m," in ", k))
  
    print(ggpaired(df , x="timepoint",y="protein", id = "patientID", line.size = .7, line.color = "RECIST",palette =  recistCol,
     title = paste0(m," in ", k), xlab="Time point", ylab = "Marker Expression") +
       theme(legend.position="right") +
      scale_x_discrete(labels=timeLab[1:2]))
      
}



```

### three time points

```{r three-time-points, echo=FALSE, fig.width= 4, fig.height= 3 }

for(i in 1:nrow(sigMarkers))
{
  # get marker
  m = sigMarkers[i,"marker"]
  # get cluster
  k = sigMarkers[i,"cluster"]
  # get expression for the cluster
  d = expr %>% filter(cluster == k)
  # create a data.frame to plot
  df = cbind(protein = d[,m], d[,c("patientID", "timepoint", "RECIST") ] )
  
  # print current pair
  print(paste0(m," in ", k))

      print(ggpaired(df , x="timepoint",y="protein", id = "patientID", line.size = .7, line.color = "RECIST",palette =  recistCol,
     title = paste0(m," in ", k), xlab="Time point", ylab = "Marker Expression") +
       theme(legend.position="right") +
      scale_x_discrete(labels=timeLab))
      
}


```

```{r echo=FALSE}
sessionInfo()
```




