---
title: "J1798 Patient table, OS, PFS, ORR, and so"
author: "Leslie Cope"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
require(kableExtra)
require(ggplot2)
require(survminer)
require(flextable)
library(binom)
library(survival)

# set root folder
rootDir = "../"

# load shared parameters and functions
source(paste0(rootDir,"parameters_and_functions.r"))

```

## Patient Characteristics


```{r dataPrep, echo=F,message=FALSE,warning=FALSE}

#===========================
# read in patient annotation 
patientTab = read.csv(file = paste0(rootDir,"input_data/table_patients.csv"), row.names = 1)
cat("The number of patients:", nrow(patientTab))
table(patientTab$RECIST)

```



```{r summarizeVars,echo=F,message=F,warning=F}
#### summarize the character and numeric variables 
numSums=sapply(colnames(patientTab)[c(1,10)],numSummary,data=patientTab)
charSums=sapply(colnames(patientTab)[c(2:9)],charSummary,data=patientTab)

### reorder differentiation values in +/- increasing order
charSums$Differentiation=charSums$Differentiation[c(3,1,2,5,4)]

## simplify a few variable names
names(charSums)=c("Gender","Race","Ethnicity","Location of Primary Tumor","Differentiation","Stage at Diagnosis","Surgical Resection","Lines of Chemo")

```


```{r flexTab1,echo=F,message=F,warning=F}
### make the patient characteristic table


### now the flex table.  
### 3 column format, variable name, mean, range for numeric
### variable name, value, count, for character 
ft=flextable(as.data.frame(t(numSums)))
for(i in 1:length(charSums)){
ft <- add_body(ft, values = setNames(modChar(names(charSums)[i]), names(as.data.frame(t(numSums)))), top = FALSE)}
lineLocs=rep(NA,length(charSums))
runSum=0
for(i in 1:(length(charSums))){
  lineLocs[i]=runSum+length(charSums[[i]])+1
runSum=lineLocs[i]  }
lineLocs=c(0,lineLocs)
ft <- hline(ft, i = lineLocs+2)
ft=autofit(ft)
ft

```



## Data Analysis

### Primary Objective (N=27 evaluable patients)

To determine the objective response rate (ORR) of entinostat plus nivolumab in patients with unresectable or metastatic pancreatic adenocarcinoma, who have progressed after one more two lines of prior therapy.


```{r ORR.27,echo=F,message=F,warning=F}
# exclude patients with toxicity
patientTab_27=patientTab[which(patientTab[,"Reason.off.Study"] != "Toxicity"),]

test=binom.test(length(grep("^Responder",patientTab_27[,"RECIST"])),nrow(patientTab_27),p=0.05,alternative="greater")
resp=binom.confint(length(grep("^Responder",patientTab_27[,"RECIST"])),nrow(patientTab_27),method=c("exact","wilson"))
orr=round(resp$mean[1],3)
orrCt=3
CI=paste0("(",round(resp$lower,3),"-",round(resp$upper,3),")")
p=round(test$p.value,3)
#names(CI)=paste0(c("exact","Wilson")," CI")
orrTab=data.frame("ORR"=orr,"exact CI"=CI[1],"p-value"=p)
orrTab.ft=flextable(orrTab)
orrTab.ft=autofit(orrTab.ft)
orrTab.ft

```


### Overall Survival (N=30)
To determine the overall survival (OS) status of subjects with unresectable or metastatic cholangiocarcinoma and unresectable or metastatic pancreatic adenocarcinoma treated with entinostat plus nivolumab at 6 months, at 1, 2 and 3 years.

#### Median Overall Survival in Months, with 95% Confidence Interval
  
  
```{r OS,echo=F,message=F,warning=F}
## make and fit survival object
survOS=Surv(time=patientTab$OS_time,event=patientTab$OS_event)
fitOS=survfit(survOS~1)


#### make and format median table
medOS=as.data.frame(median(survOS))
sumMed=function(x){ #x = surv object
medOS=as.data.frame(median(x))
if(is.na(medOS$upper)) medOS$upper=Inf
Median=round(medOS$q,3)
CI=paste0("(",round(medOS$lower,3),"-",round(medOS$upper,3),")")
return(data.frame("Median Survival"=Median,"CI"=CI, check.names = F))
}

medTab=sumMed(survOS)
medTab.ft=flextable(medTab)
medTab.ft=autofit(medTab.ft)
medTab.ft


```



#### K-M plot of overall survival, with at risk table
 

```{r OS.KM, echo=F, fig.height= 5}
## make and format plot with at risk table
plotOS=ggsurvplot(fitOS,data=survOS,risk.table = TRUE,legend="none",title="Overall Survival") + xlab("Time (Months)")

plotOS$plot = plotOS$plot +
  # add horizontal line at median
  geom_hline(yintercept = .5, linetype="dashed", color = 'grey25',) +
  # add numbers for median survival and CI
  annotate("text", x = 25, y = 0.65,size = 3.5, 
           label = paste0("Median Survival\n",medTab[1]," [95% CI ",medTab[2],"]"))

plotOS
```



### Progression Free Survival (N=30)
To assess the progression free survival (PFS) status of patients with unresectable or metastatic cholangiocarcinoma and unresectable or metastatic pancreatic adenocarcinoma treated with entinostat plus nivolumab at 6, 12 and 24 months.

#### Median Progression Free Survival in Months, with 95% Confidence Interval
```{r pfsSetup, echo=F,message=F,warning=F}
## make and fit survival object

survPFS=Surv(time=patientTab$PFS_time,event=patientTab$PFS_event)
fitPFS=survfit(survPFS~1)


#### make and format median table
medPFS=as.data.frame(median(survPFS))


medTab=sumMed(survPFS)
medTabPFS.ft=flextable(medTab)
medTabPFS.ft=autofit(medTabPFS.ft)
medTabPFS.ft
```


#### K-M plot of progression free survival, with at risk table
 

```{r PFS.KM, echo=F, fig.height=5}
## make and format plot with at risk table
plotPFS=ggsurvplot(fitPFS,data=survPFS,risk.table = TRUE,legend="none",title="Progression Free Survival")+
  xlab("Time (Months)") 

plotPFS$plot = plotPFS$plot +
  # add horizontal line at median
  geom_hline(yintercept = .5, linetype="dashed", color = 'grey25',) +
  # add numbers for median survival and CI
  annotate("text", x = 15, y = 0.65,size = 3.5, 
           label = paste0("Median Survival\n",medTab[1]," [95% CI ",medTab[2],"]"))

plotPFS
```


