---
title: "J1798 Patient table, OS, PFS, ORR, and so"
author: "Leslie Cope"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
require(kableExtra)
require(flextable)
require(ggplot2)
require(survminer)
require(flextable)
library(binom)
library(survival)

# set root folder
rootDir = "../"

# load shared parameters and functions
source(paste0(rootDir,"parameters_and_functions.r"))

```

## Patient Characteristics


```{r dataPrep, echo=F,message=FALSE,warning=FALSE}

#===========================
# read in patient annotation 
patientTab = read.csv(file = paste0(rootDir,"input_data/table_patients.csv"), row.names = 1)
cat("The number of patients:", nrow(patientTab))
table(patientTab$RECIST)

```



```{r summarizeVars,echo=F,message=F,warning=F}
#### summarize the character and numeric variables 
numSums=sapply(colnames(patientTab)[c(1,10)],numSummary,data=patientTab)
charSums=sapply(colnames(patientTab)[c(2:9)],charSummary,data=patientTab)

### reorder differentiation values in +/- increasing order
charSums$Differentiation=charSums$Differentiation[c(3,1,2,5,4)]

## simplify a few variable names
names(charSums)=c("Gender","Race","Ethnicity","Location of Primary Tumor","Differentiation","Stage at Diagnosis","Surgical Resection","Lines of Chemo")

```


```{r flexTab1,echo=F,message=F,warning=F}
### make the patient characteristic table


### now the flex table.  
### 3 column format, variable name, mean, range for numeric
### variable name, value, count, for character 
ft=flextable(as.data.frame(t(numSums)))
for(i in 1:length(charSums)){
ft <- add_body(ft, values = setNames(modChar(names(charSums)[i]), names(as.data.frame(t(numSums)))), top = FALSE)}
lineLocs=rep(NA,length(charSums))
runSum=0
for(i in 1:(length(charSums))){
  lineLocs[i]=runSum+length(charSums[[i]])+1
runSum=lineLocs[i]  }
lineLocs=c(0,lineLocs)
ft <- hline(ft, i = lineLocs+2)
ft=autofit(ft)
ft

```



## Data Analysis

### Primary Objective (N=27 evaluable patients)

To determine the objective response rate (ORR) of entinostat plus nivolumab in patients with unresectable or metastatic cholangiocarcinoma and unresectable or metastatic pancreatic adenocarcinoma, who have progressed after one more two lines of prior therapy.


```{r ORR.27,echo=F,message=F,warning=F}
patientTab_27=patientTab[which(!is.na(patientTab[,"BestResponse"])),]

test=binom.test(length(grep("R",patientTab_27[,"BestResponse"])),nrow(patientTab_27),p=0.05,alternative="greater")
resp=binom.confint(length(grep("R",patientTab_27[,"BestResponse"])),nrow(patientTab_27),method=c("exact","wilson"))
orr=round(resp$mean[1],3)
orrCt=3
CI=paste0("(",round(resp$lower,3),"-",round(resp$upper,3),")")
p=round(test$p.value,3)
#names(CI)=paste0(c("exact","Wilson")," CI")
orrCt=table(patientTab_27[,"BestResponse"])
orrTab=data.frame("PR"=orrCt["PR"],"SD"=orrCt["SD"],"PD"=orrCt["PD"],"ORR"=orr,"exact CI"=CI[1],"Wilson CI"=CI[2],"p-value"=p)
orrTab.ft=flextable(orrTab)
orrTab.ft=autofit(orrTab.ft)
orrTab.ft

```

### Secondary Objectives
### Disease Control Rate



```{r DCR.27,echo=F,message=F,warning=F, eval = F}

resp.dcr=binom.confint(sum(patientTab_27[,"BestResponse"]!="PD"),nrow(patientTab_27),method=c("exact","wilson"))
orr.dcr=round(resp.dcr$mean[1],3)
orrCt.dcr=5
CI.dcr=paste0("(",round(resp.dcr$lower,3),"-",round(resp.dcr$upper,3),")")

#names(CI)=paste0(c("exact","Wilson")," CI")
orrCt.dcr=table(patientTab_27[,"BestResponse"])
orrTab.dcr=data.frame("PR"=orrCt.dcr["PR"],"SD"=orrCt["SD"],"PD"=orrCt["PD"],"DCR"=orr.dcr,"exact CI"=CI.dcr[1],"Wilson CI"=CI.dcr[2])
orrTab.dcr.ft=flextable(orrTab.dcr)
orrTab.dcr.ft=autofit(orrTab.dcr.ft)
orrTab.dcr.ft

```

### Overall Survival (N=30)
To determine the overall survival (OS) status of subjects with unresectable or metastatic cholangiocarcinoma and unresectable or metastatic pancreatic adenocarcinoma treated with entinostat plus nivolumab at 6 months, at 1, 2 and 3 years.

#### Median Overall Survival in Months, with 95% Confidence Interval
  
  
```{r OS,echo=F,message=F,warning=F}
## make and fit survival object
survOS=Surv(time=patientTab$OS_time,event=patientTab$OS_event)
fitOS=survfit(survOS~1)


#### make and format median table
medOS=as.data.frame(median(survOS))
sumMed=function(x){ #x = surv object
medOS=as.data.frame(median(x))
if(is.na(medOS$upper)) medOS$upper=Inf
Median=round(medOS$q,3)
CI=paste0("(",round(medOS$lower,3),"-",round(medOS$upper,3),")")
return(data.frame("Median Survival"=Median,"CI"=CI))
}

medTab=sumMed(survOS)
medTab.ft=flextable(medTab)
medTab.ft=autofit(medTab.ft)
medTab.ft


```


#### Overall survival rates at 6, 12, 24, and 36 months, with 95% Confidence Intervals

```{r survOS,echo=F,message=F,warning=F}
## make and format time table
osSum=summary(fitOS, times=c(6,12,24,36),extend=T)
sumSurv=function(x){
Time=paste0(x$time," Month Survival")
Survival=round(x$surv,3)
CI=paste0("(",round(x$lower,3),"-",round(x$upper,3),")")
ans=data.frame(Time,Survival,CI)
return(ans)
}

osTab=sumSurv(osSum)
osTab.ft=flextable(osTab)
osTab.ft=autofit(osTab.ft)
osTab.ft
```


#### K-M plot of overall survival, with at risk table
 

```{r OS.KM, echo=F, fig.height= 5}
## make and format plot with at risk table
plotOS=ggsurvplot(fitOS,data=survOS,risk.table = TRUE,legend="none",title="Overall Survival") + xlab("Time (Months)") 

plotOS

```



### Progression Free Survival (N=30)
To assess the progression free survival (PFS) status of patients with unresectable or metastatic cholangiocarcinoma and unresectable or metastatic pancreatic adenocarcinoma treated with entinostat plus nivolumab at 6, 12 and 24 months.

#### Median Progression Free Survival in Months, with 95% Confidence Interval
```{r pfsSetup, echo=F,message=F,warning=F}
## make and fit survival object

survPFS=Surv(time=patientTab$PFS_time,event=patientTab$PFS_event)
fitPFS=survfit(survPFS~1)


#### make and format median table
medPFS=as.data.frame(median(survPFS))


medTab=sumMed(survPFS)
medTabPFS.ft=flextable(medTab)
medTabPFS.ft=autofit(medTabPFS.ft)
medTabPFS.ft
```

#### Progression Free survival rates at 6, 12, and 24* months, with 95% Confidence Intervals
*All patients progressed by 24mo. 
```{r survPFS,echo=F,message=F,warning=F}
## make and format time table
pfsSum=summary(fitPFS, times=c(6,12,24),extend=T)

pfsTab=sumSurv(pfsSum)
pfsTab.ft=flextable(pfsTab)
pfsTab.ft=autofit(pfsTab.ft)
pfsTab.ft
```


#### K-M plot of progression free survival, with at risk table
 

```{r PFS.KM, echo=F, fig.height=5}
## make and format plot with at risk table
plotPFS=ggsurvplot(fitPFS,data=survPFS,risk.table = TRUE,legend="none",title="Progression Free Survival")+
  xlab("Time (Months)") 
plotPFS

```


### Duration of Response
To assess the duration of response among subjects who demonstrate an objective response to treatment with entinostat in combination with nivolumab.


```{r duration, echo=F, eval = F}
bestRespDate=data[,77]
duration=(datePFS-bestRespDate)*12/365
durV=as.numeric(round(sort(duration),1))
durV=paste0(durV," Months")
dur=data.frame(durV)
colnames(dur)="Duration of response"
ft.dur=flextable(dur)
ft.dur=autofit(ft.dur)
ft.dur
save.image()

```


#### Analysis of Secondary Objectives

Median time to event will be calculated and reported with confidence intervals.

OS curves, OS medians with 95% CIs, and OS rates at 6, 12, 24 and 36 months with 95% CIs will be estimated using Kaplan-Meier methodology

PFS curves, PFS medians with 95% CIs, and PFS rates at 6, 12 and 24 months with 95% CIs will be estimated using Kaplan-Meier methodology.


Duration of response (DOR) will be calculated for subjects who achieve a best overall response of CR or PR with entinostat in combination with nivolumab. For such subjects, duration of response is defined as the number of weeks from the start date of PR or CR (whichever response is recorded first) and subsequently confirmed to the first date that recurrent or progressive disease or death is documented. DOR will be summarized descriptively.

